---
title: "2025 Data Fest"
subtitle: "Introduction to R, Text Mining, and Sentiment Analysis"
author: "[Randi Bolt](https://randibolt.com/)"
date: "9/25/2025"
format:
  revealjs:
    theme: serif
    logo: img/bolt.png
    footer: "2025-datafest.netlify.app"
    incremental: true 
---

## About Me

I’m the Data Equity and Innovation Supervisor for **[Paid Leave Oregon](https://paidleave.oregon.gov/)**, where I lead a team of data analysts in transforming complex data into insights.

<br>

Outside of work, I explore data through various personal projects that incorporate analytics, visualization, and storytelling. Check out my [blogs](https://2024-rbolt.netlify.app/projects/technical-blogs), [dashboards](https://randibolt.com/dashboards), and [talks](https://randibolt.com/talks) to see more.

## Agenda

* Introduction to R & R Studio


## Introduction to R & R Studio 

{{< video src="https://www.youtube.com/watch?v=wIft-t-MQuE" width="100%" height="80%" >}}

## What is R

R is a statistical programming language that's incredibly powerful for working with data. 

<br>

Unlike Python which is built around _objects_, R is based on _functions_. For our purposes, that means we'll be calling functions (like $F(x)=Y$) to transform our data rather than attaching properties to objects. 


## Benefits of R

* R is open source and freely available. 

* R has an extensive and coherent set of tools for statistical analysis. 

* R has an extensive and highly flexible graphical facility capable of production publication quality figures. 

* R has an extensive support network with numerous online and freely available documents. 

* R has an expanding set of freely available 'packages' to extend R's capabilities. [^1]

[^1]: [Intro to R](https://intro2r.com/chap1.html)

## What is R Studio 

R Stuido is an **I**ntegrated **D**evelopment **E**nvironment (IDE), similar to VS Code or Sublime. 

<br>

R Studio provides a more user-friendly interface, incorporating the R Console, a script editor and other useful functionality (like R markdown and GitHub integration). You can find more information about [RStudio here](). [^2]

[^2]: [Intro to R](https://intro2r.com/chap1.html)

## Live Demo

* Landscape (Console, help, viewer, environment, files)

* Global Options 

## Packages 

Most of the work we'll do relies on packages, which are basically toolkits. To use one, you will first need to install it with a base R function `install.packages()`: 


```{r}
#| echo: true
#| eval: false
install.packages("package_name")
```

After installing a package, you can load it into your current session with `library()`:

```{r}
#| echo: true
#| eval: false
library("package_name")
```

## Tips

When in doubt about what a functions does, or what is in a package, you can type `?function_name()` or `?package_name` in the R Studio console to open the description page in the Help tab. 

## CRAN

CRAN is the Comprehensive R Archive Network. It’s where R packages are stored, tested, and shared with the community. If you install a package in R, you’re usually pulling it from CRAN.

## Taylor Swift Data

{{< video src="https://www.youtube.com/watch?v=e-ORhEE9VVg" width="100%" height="80%" >}}

## Taylor Package

The Taylor package is a comprehensive resource for data on Taylor Swift songs. Data comes from 'Genius' (lyrics) and 'Spotify' (song characteristics). 

**Useful links**: [taylor](https://taylor.wjakethompson.com/), [taylor repo](https://github.com/wjakethompson/taylor)


### Install & Load Data

```{r}
#| warning: false
#| error: false
install.packages("taylor", repos="http://cran.us.r-project.org")
library("taylor")
```

```{r}
#| echo: true
#| eval: false
install.packages("taylor")
library("taylor")
```

## Taylor Data

There are three main data sets: `taylor_album_songs`

```{r}
#| echo: true
head(taylor_album_songs)
```

## Reactable

This package creates a data table with sorting and pagination. The default table is an HTML widget that can be used in RMD and Shiny applications, or viewed from an R console. 

```{r}
#| warning: false
#| error: false
install.packages("reactable", repos="http://cran.us.r-project.org")
library("reactable")
```

```{r}
#| echo: true
#| eval: false
install.packages("reactable")
library("reactable")
```

## Reactable Taylor Data

```{r}
#| echo: true
reactable(
  taylor_album_songs,
  wrap = FALSE,
  defaultPageSize = 4,
  defaultColDef = colDef(minWidth = 300))
```
